Can machine learning correct microwave humidity radiances for the influence of clouds?

### 2020-11-24 Article/ici_submm_cloudcorr.tex
Updated with Dave's comments, New title

\title{Cloud correction of humidity radiances with case-specific uncertainty 
	estimate}

@@

\runningtitle{Cloud correction with case specific uncertainty estimation}

@@

Another motivation necessitating the need for cloud clearing is usage of 183\,GHz channels in 
numerical weather prediction (NWP). Usage of passive microwave data by
all-sky assimilation in global NWP is growing \citep{geer2017growing},
but 183\,GHz data are still mainly used in a clear-sky fashion
\citep{geer2018all}. The latter is particularly true in NWP of regional scope
\citep{gustafsson2018survey}, with clear-sky assimilation of 183\,GHz radiances still commonplace. Regardless, both clear-sky and all-sky assimilation require identification of cloud affected observations, either to screen out these observations or to assign an appropriate observation error. The most commonly used cloud filtering techniques are the ``Scattering index'' \citep{geer2015scatteringindex} and the ``observation minus background'' (O$-$B). The first one is based on brightness temperature differences between 89 and 150\,GHz. In the second one, the forecast model is used to obtain an estimate of the expected clear-sky value and the observation is rejected if the deviation exceeds some threshold \citep{English1999clouddetection}.

@@

\subsubsection{QRNN-single applied to MWHS-2 channel 14} 
\begin{figure}[t]
	\centering
	\includegraphics[width = 70mm]{Figures/fig02.pdf} 
	\caption{The distribution of point estimates (``Predicted'') obtained from QRNN-single 89+150\,GHz for MWHS-2 channel 14. The corresponding distributions for all-sky and clear-sky simulations are also shown.}
	\label{fig:distribution_predicted_mwhs14}	
\end{figure}
\begin{figure}[t]
	\centering
	\includegraphics[width = 70mm]{Figures/fig03.pdf} 
	\caption{The error distribution for deviations of point estimates to NFCS simulations. The results are from QRNN-single experiment 89+150\,GHz and MWHS-2 channel 14. Noise is also plotted for reference. The label ``All-sky'' represents the all-sky simulations, ``Predicted (All)'' denotes the predicted point estimates. The error distribution achieved with scattering index (SI) filtering is also shown (Filtered (SI)). }
	\label{fig:error_distribution_mwhs14}	
\end{figure}
\begin{table*}[t]
	\caption{The error statistics for deviations of point estimates to NFCS simulations. Results are for different QRNN experiments for MWHS-2 channel 14 (see Sect.~\ref{sec:QRNN_expt_MWHS}). The statistics for all-sky and clear-sky simulations are also provided. The label ``All'' denotes the entire dataset of predicted point estimates, while ``Pred. (5\,K)'' refers to the predicted point estimates but where cases with cloud correction greater than 5\,K are excluded. The last two columns show the statistics obtained after filtering cloudy cases according to scattering index (SI) and scheme by \citet{buehler:aclou:07} (B183). Bias, MAE, STD are in K, skewness is dimensionless.}
	\label{tab:error_statistics_mwhs_14}
	\setlength{\tabcolsep}{4pt}
	\begin{tabular}{lrr|rr|rr|rr|rr|rr}
		\tophline
		&\multicolumn{2}{c|}{Simulations}& \multicolumn{6}{c|}{QRNN-single} & \multicolumn{2}{c|}{QRNN-all} & \multicolumn{2}{c}{Pure filtering}\\
		\cline{2-13}
		%		\hline
		&   Clear-sky &   All-sky &  \multicolumn{2}{c|}{89+150\,GHz} & \multicolumn{2}{c|}{89+150+118\,GHz} & \multicolumn{2}{c|}{150\,GHz} & \multicolumn{2}{c|}{89+150+183\,GHz}& SI & B183\\		
		&			   &			& All & Pred. (5\,K) & All & Pred. (5\,K) & All & Pred. (5\,K)  & All & Pred. (5\,K)&&\\
		\middlehline
Bias     &  0.00 & $-$0.84 &$-$0.10 &$-$0.09 &$-$0.05 &$-$0.04 &$-$0.11 &$-$0.10 &$-$0.10 &$-$0.09 & 0.24  &$-$0.52\\
MAE      &  0.80 &   1.45 &  0.89 &  0.87 &  0.90 &  0.88 &  0.90 &  0.88 &  0.62 &  0.60 & 0.92  & 1.15\\
STD      &  1.00 &   3.73 &  1.20 &  1.16 &  1.21 &  1.16 &  1.22 &  1.18 &  0.91 &  0.85 & 1.26  & 1.86\\
Skewness &$-$0.02 &$-$12.72 &$-$1.17 &$-$1.00 &$-$1.06 &$-$0.90 &$-$1.16 &$-$1.04 &$-$1.78 &$-$1.71 &-1.95  &$-$3.45\\
Rejection& -	 &- 	  &-	  & 3.3\,\% &- 	  & 3.4\,\% &-     & 3.2\,\% &-     & 3.3\,\% & 28.8\,\%& 3.5\,\%\\
\bottomhline
	\end{tabular}
\end{table*}

Posterior distributions of $\ynfcs$ obtained from experiment 89+150\,GHz are similar to the ones shown in Fig.~\ref{fig:posterior_distribution_I1V} and the distribution of point estimates is displayed in Fig.~\ref{fig:distribution_predicted_mwhs14}. The predicted values are able to correct most of the low brightness temperature cases, and overall a good match with the NFCS simulations is observed. However, QRNN is unable to predict the lowest clear-sky brightness temperatures, and cases with brightness temperatures around 260\,K, occur too frequently. The deviations of point estimates to NFCS simulations are shown in Fig.~\ref{fig:error_distribution_mwhs14}. The large negative deviations are removed (blue curve), but residual cloud impact is evident in the negative tail. Most of these residual cases have departures less than 10\,K. The appearance of a small positive tail also indicates overestimation in few cases. The corresponding error statistics (see Sect.~\ref{sec:validation}) are provided in  Table~\ref{tab:error_statistics_mwhs_14}. In the uncorrected all-sky simulations, the negative departures due to cloud impact lead to a large negative skewness ($-$12.72) and high bias ($-$0.84\,K). QRNN trained with 89 and 150\,GHz successfully corrects a major portion of the cloud affected cases, and the bias is reduced to $-$0.10\,K. However, negative skewness indicates presence of  uncorrected departures. Including 118\,GHz in the training gives a small improvement compared to 89+150 GHz alone. The error distribution has lower bias and is more symmetric, but the MAE and standard deviation remain unaltered. This indicates that the information from 118\,GHz channels can be beneficial in predicting few cases correctly, but the overall performance is not exceptionally different from 89+150\,GHz. Similar is the case with using only 150\,GHz---the differences between the three experiments are negligible.

@@

\subsubsection{Comparison of QRNN-all and QRNN-single}
%
%f
\begin{figure*}[t]
	\centering
	\includegraphics[height = 55mm]{Figures/fig04.pdf} 
	\caption{ The triangular error correlation matrix obtained from (a) QRNN-single and (b) QRNN-all for MWHS-2. The label ``Clear'' represents cases with cloud impact less than 2\,K.}
	\label{fig:correlations_mwhs}	
\end{figure*}
To assess the differences between the capabilities of QRNN-all against QRNN-single, we  compare the error statistics obtained for 89+150\,GHz and 89+150+183\,GHz for channel 14.  Table~\ref{tab:error_statistics_mwhs_14} also shows the error statistics for the experiment QRNN-all. In comparison to QRNN-single, we obtain almost similar error bias with QRNN-all, but the MAE and standard deviation reduce by almost 30\,\% and 24\,\% respectively. However, the negative tail becomes more prominent. The low standard deviation, but strong negative tail indicates that the narrow spread is a consequence of correction of noise in clear-sky cases. Since a majority of the cases are clear-sky, their impact dominates the whole statistics. In order to probe the positive effect of QRNN-all, we also estimate the errors for all cases with cloud correction greater than 5\,K (table not shown). For such cases,  QRNN-all has slightly better accuracy than QRNN-single. The bias in QRNN-all and standard deviation is $-$0.53\,K and 1.93\,K, respectively, in comparison to $-$0.61\,K and 2.04\,K as observed in QRNN-single. Thus, the concurrent use of all 183\,GHz channels can provide additional information on cloud structures to QRNN.

@@

The partial performance of QRNN is due to incomplete orthogonal information to 183\,GHz channels. The weighting functions of window channels 89 and 150\,GHz, and the two 118\,GHz channels peak in the lower troposphere. Among these four channels, 150\,GHz has the highest peaking function around 4\,km \citep{chen2020mwhs}. These channels can only provide coverage to the humidity channels in, lower and mid troposphere; nonetheless, the 183\,GHz channels are sensitive to hydrometeor content up to 10\,km. Due to missing complementary information from other channels in the upper troposphere, QRNN fails at predicting these cases accurately. Such cases are mostly associated with thin cirrus clouds, which have very small influence at 89 and 150\,GHz. Without additional information from other channels we cannot expect QRNN to perform better. Among the other available channels, the overlapping weighting functions of 183\,GHz can provide auxiliary information to train QRNN. Results show that these channels indeed help in improving the training, yet the non-orthogonal information introduces highly correlated observational errors. The correlations for cloudy observations are not surprising as the cloud amount for different channels depends upon each other in a systematic way. However for clear-sky observations, the correlations should preferably be close to zero. This is observed to be true with QRNN-single, but QRNN-all fails at preserving the noise stochasticity. In the absence of hydrometeor impact, all 183\,GHz channels provide same information to the learning model, introducing redundancy. Redundant patterns in machine learning models often have undesirable effect on the predictive performance. In our application, redundant information does not affect the prediction accuracy, but introduces highly correlated observational errors between channels. Correlated errors are also undesirable for DA systems, but in future, if the operational centres progress with approaches dealing with correlated observation errors, for example as currently done for ATMS at ECMWF \citep{Weston2018eumATMS}, concurrent use of 89, 150 and 183\,GHz channels would give the best cloud correction performance.

@@

Another advantage of QRNN is the estimation of case-specific uncertainties. That is, the predictions over chosen quantiles quantify the underlying uncertainty of the particular case, and not just represent some ensemble mean error. This has the consequence that a DA system can assign a proper weighting of each individual QRNN prediction.
The analysis of QRNN predicted errors and calibration plots confirmed that QRNN is successful in providing
well calibrated probabilistic predictions also in practice, except for few cases associated with high error. Poorly calibrated predictions are a consequence of cases with different underlying a priori distribution. Predictions with high error originate from cases which occur infrequently or cases with no orthogonal information. Such cases have a significantly distinct a priori distribution as compared to clear cases, which dominate the training dataset, and influence the a priori information. The distribution of rare cases can be improved by increasing the training dataset size, but lack of orthogonal information can only be balanced by including additional training inputs, e.g. brightness temperatures for other channels.

The symmetric and low spread error distribution with uncertainty estimates is also an important result from the DA perspective. Most of the existing cloud filtering schemes work well only at removing cases with high cloud impact, and as a consequence, the error distributions are highly skewed. To use these observations correctly, DA schemes often inflate their asymmetric error distributions at the cost of artificially suppressing the observational impact. However, the symmetric error distributions obtained from QRNN allow effective utilization of almost complete data without the need for artificial error inflation. In fact for DA, filtering based on correction threshold would be needless as cases with low accuracy shall inevitably get inhibited due to high uncertainty.

@@

In this study, a methodology based on quantile regression neural network (QRNN) is used for identifying and correcting the cloud contamination in operational humidity channels. QRNN is a neural network which trains on all-sky brightness temperatures from  channels containing orthogonal information to humidity channels, to estimate the noise free clear-sky (NFCS) brightness temperatures. The output is the posterior distribution of predictions over different quantiles. QRNN is a channel specific approach, or in other words, QRNN is trained separately for each channel, and the cloud correction for each is independent of other channels.

@@

The accuracy of QRNN is affected by one single limitation, that is presence of cases which belong to a different a priori distribution than the one QRNN is trained on. Such cases can be either the ones which occur infrequently, or those which lack any orthogonal information. The representation of the former can be improved to represent reality by simply choosing larger training datasets. However for the latter, improving their a priori distribution is not trivial. Correcting hydrometeor impact for cases without complete complementary information is a challenge for QRNN and always limits its performance.

>>>>

\title{Can machine learning correct microwave humidity radiances for the
	influence of clouds?}

@@

\runningtitle{Correcting cloud affected microwave radiances}

@@

Another motivation necessitating the need for cloud filtering is usage of 183\,GHz channels in 
numerical weather prediction (NWP). Usage of passive microwave data by
all-sky assimilation in global NWP is growing \citep{geer2017growing},
but 183\,GHz data are still mainly used in a clear-sky fashion
\citep{geer2018all}. The latter is particularly true in NWP of regional scope
\citep{gustafsson2018survey}, with clear-sky assimilation of 183\,GHz radiances still commonplace. Regardless, both clear-sky and all-sky assimilation require identification of cloud affected observations, either to screen out these observations or to assign an appropriate observation error. The most commonly used cloud filtering techniques are the ``Scattering index'' \citep{geer2015scatteringindex} and the ``observation minus background'' (O$-$B). The first one is based on brightness temperature differences between 89 and 150\,GHz. In the second one, the forecast model is used to obtain an estimate of the expected clear-sky value and the observation is rejected if the deviation exceeds some threshold \citep{English1999clouddetection}.

@@

\subsubsection{QRNN-single applied to MWHS-2 channel 14} 
\begin{figure}[t]
	\centering
	\includegraphics[width = 70mm]{Figures/fig02.pdf} 
	\caption{The distribution of point estimates (``Predicted'') obtained from QRNN-single 89+150\,GHz for MWHS-2 channel 14. The corresponding distributions for all-sky and clear-sky simulations are also shown.}
	\label{fig:distribution_predicted_mwhs14}	
\end{figure}
\begin{figure}[t]
	\centering
	\includegraphics[width = 70mm]{Figures/fig03.pdf} 
	\caption{The error distribution for deviations of point estimates from NFCS simulations. The results are from QRNN-single experiment 89+150\,GHz and MWHS-2 channel 14. Noise is also plotted for reference. The label ``All-sky'' represents the all-sky simulations, ``Predicted (All)'' denotes the predicted point estimates. The error distribution achieved with scattering index (SI) filtering is also shown (Filtered (SI)). }
	\label{fig:error_distribution_mwhs14}	
\end{figure}
\begin{table*}[t]
	\caption{The error statistics for deviations of point estimates from NFCS simulations. Results are for different QRNN experiments for MWHS-2 channel 14 (see Sect.~\ref{sec:QRNN_expt_MWHS}). The statistics for all-sky and clear-sky simulations are also provided. The label ``All'' denotes the entire dataset of predicted point estimates, while ``Pred. (5\,K)'' refers to the predicted point estimates but where cases with cloud correction greater than 5\,K are excluded. The last two columns show the statistics obtained after filtering cloudy cases according to scattering index (SI) and scheme by \citet{buehler:aclou:07} (B183). Bias, MAE, STD are in K, skewness is dimensionless.}
	\label{tab:error_statistics_mwhs_14}
	\setlength{\tabcolsep}{4pt}
	\begin{tabular}{lrr|rr|rr|rr|rr|rr}
		\tophline
		&\multicolumn{2}{c|}{Simulations}& \multicolumn{6}{c|}{QRNN-single} & \multicolumn{2}{c|}{QRNN-all} & \multicolumn{2}{c}{Pure filtering}\\
		\cline{2-13}
		%		\hline
		&   Clear-sky &   All-sky &  \multicolumn{2}{c|}{89+150\,GHz} & \multicolumn{2}{c|}{89+150+118\,GHz} & \multicolumn{2}{c|}{150\,GHz} & \multicolumn{2}{c|}{89+150+183\,GHz}& SI & B183\\		
		&			   &			& All & Pred. (5\,K) & All & Pred. (5\,K) & All & Pred. (5\,K)  & All & Pred. (5\,K)&&\\
		\middlehline
Bias     &  0.00 & $-$0.84 &$-$0.10 &$-$0.09 &$-$0.05 &$-$0.04 &$-$0.11 &$-$0.10 &$-$0.10 &$-$0.09 & 0.24  &$-$0.52\\
MAE      &  0.80 &   1.45 &  0.89 &  0.87 &  0.90 &  0.88 &  0.90 &  0.88 &  0.62 &  0.60 & 0.92  & 1.15\\
STD      &  1.00 &   3.73 &  1.20 &  1.16 &  1.21 &  1.16 &  1.22 &  1.18 &  0.91 &  0.85 & 1.26  & 1.86\\
Skewness &$-$0.02 &$-$12.72 &$-$1.17 &$-$1.00 &$-$1.06 &$-$0.90 &$-$1.16 &$-$1.04 &$-$1.78 &$-$1.71 &-1.95  &$-$3.45\\
Rejection& -	 &- 	  &-	  & 3.3\,\% &- 	  & 3.4\,\% &-     & 3.2\,\% &-     & 3.3\,\% & 28.8\,\%& 3.5\,\%\\
\bottomhline
	\end{tabular}
\end{table*}

Posterior distributions of $\ynfcs$ obtained from experiment 89+150\,GHz are similar to the ones shown in Fig.~\ref{fig:posterior_distribution_I1V} and the distribution of point estimates is displayed in Fig.~\ref{fig:distribution_predicted_mwhs14}. The predicted values are able to correct most of the low brightness temperature cases, and overall a good match with the NFCS simulations is observed. However, QRNN is unable to predict the lowest clear-sky brightness temperatures, and cases with brightness temperatures around 260\,K, occur too frequently. The deviations of point estimates from NFCS simulations are shown in Fig.~\ref{fig:error_distribution_mwhs14}. The large negative deviations are removed (blue curve), but residual cloud impact is evident in the negative tail. Most of these residual cases have departures less than 10\,K. The appearance of a small positive tail also indicates overestimation in few cases. The corresponding error statistics (see Sect.~\ref{sec:validation}) are provided in  Table~\ref{tab:error_statistics_mwhs_14}. In the uncorrected all-sky simulations, the negative departures due to cloud impact lead to a large negative skewness ($-$12.72) and high bias ($-$0.84\,K). QRNN trained with 89 and 150\,GHz successfully corrects a major portion of the cloud affected cases, and the bias is reduced to $-$0.10\,K. However, negative skewness indicates presence of  uncorrected departures. Including 118\,GHz in the training gives a small improvement compared to 89+150 GHz alone. The error distribution has lower bias and is more symmetric, but the MAE and standard deviation remain unaltered. This indicates that the information from 118\,GHz channels can be beneficial in predicting few cases correctly, but the overall performance is not exceptionally different from 89+150\,GHz. Similar is the case with using only 150\,GHz---the differences between the three experiments are negligible.

@@

\subsubsection{Comparison of QRNN-all and QRNN-single}
%
%f
\begin{figure*}[t]
	\centering
	\includegraphics[width=0.65\textwidth]{Figures/fig04.pdf} 
	\caption{ The triangular error correlation matrix obtained from (a) QRNN-single and (b) QRNN-all for MWHS-2. The label ``Clear'' represents cases with cloud impact less than 2\,K.}
	\label{fig:correlations_mwhs}	
\end{figure*}
To assess the differences between the capabilities of QRNN-all against QRNN-single, we  compare the error statistics obtained for 89+150\,GHz and 89+150+183\,GHz for channel 14.  Table~\ref{tab:error_statistics_mwhs_14} also shows the error statistics for the experiment QRNN-all. In comparison to QRNN-single, we obtain almost similar error bias with QRNN-all, but the MAE and standard deviation reduce by almost 30\,\% and 24\,\% respectively. However, the negative tail becomes more prominent. The low standard deviation, but strong negative tail indicates that the narrow spread is a consequence of correction of noise in clear-sky cases. Since a majority of the cases are clear-sky, their impact dominates the whole statistics. In order to probe the positive effect of QRNN-all, we also estimate the errors for all cases with cloud correction greater than 5\,K (table not shown). For such cases,  QRNN-all has slightly better accuracy than QRNN-single. The bias in QRNN-all and standard deviation is $-$0.53\,K and 1.93\,K, respectively, in comparison to $-$0.61\,K and 2.04\,K as observed in QRNN-single. Thus, the concurrent use of all 183\,GHz channels can provide additional information on cloud structures to QRNN.

@@

The partial performance of QRNN is due to incomplete complementary information to 183\,GHz channels. The weighting functions of window channels 89 and 150\,GHz, and the two 118\,GHz channels peak in the lower troposphere. Among these four channels, 150\,GHz has the highest peaking function around 4\,km \citep{chen2020mwhs}. These channels can only provide coverage to the humidity channels in, lower and mid troposphere; nonetheless, the 183\,GHz channels are sensitive to hydrometeor content up to 10\,km. Due to missing complementary information from other channels in the upper troposphere, QRNN fails at predicting these cases accurately. Such cases are mostly associated with thin cirrus clouds, which have very small influence at 89 and 150\,GHz. Without additional information from other channels we cannot expect QRNN to perform better. Among the other available channels, the overlapping weighting functions of 183\,GHz can provide auxiliary information to train QRNN, but such information would be not be completely orthogonal or statistically uncorrelated. Results show that these channels indeed help in improving the training, yet non-orthogonality introduces highly correlated observational errors. The correlations for cloudy observations are not surprising as the cloud amount for different channels depends upon each other in a systematic way. However for clear-sky observations, the correlations should preferably be close to zero. This is observed to be true with QRNN-single, but QRNN-all fails at preserving the noise stochasticity. In the absence of hydrometeor impact, all 183\,GHz channels provide same information to the learning model, introducing redundancy. Redundant patterns in machine learning models often have undesirable effect on the predictive performance. In our application, redundant information does not affect the prediction accuracy, but introduces highly correlated observational errors between channels. Correlated errors are also undesirable for DA systems, but in future, if the operational centres progress with approaches dealing with correlated observation errors, for example as currently done for ATMS at ECMWF \citep{Weston2018eumATMS}, concurrent use of 89, 150 and 183\,GHz channels would give the best cloud correction performance.

@@

Another advantage of QRNN is the estimation of case-specific uncertainties. That is, the predictions over chosen quantiles quantify the underlying uncertainty of the particular case, and not just represent some ensemble mean error. This has the consequence that a DA system can assign a proper weighting of each individual QRNN prediction. The analysis of QRNN predicted errors and calibration plots confirmed that QRNN is successful in providing well calibrated probabilistic predictions also in practice, except for few cases associated with high error. Poorly calibrated predictions are a consequence of outliers which are not well represented in the training dataset. The a priori distribution of the training dataset is dominated by clear-sky cases, however, the cloudy cases which occur infrequently or lack independent complementary information cannot be represented by same a priori distribution. The distribution of rare cases can be improved by increasing the training dataset size, but lack of complementary information can only be balanced by including additional training inputs, e.g. brightness temperatures for other channels.

The symmetric and low spread error distribution with uncertainty estimates is also an important result from the DA perspective. Most of the existing cloud filtering schemes work well only at removing cases with high cloud impact, and as a consequence, the error distributions are highly skewed. To use these observations correctly, DA schemes often inflate their assigned observation errors at the cost of artificially suppressing the observational impact. However, the symmetric error distributions obtained from QRNN allow effective utilization of almost complete data without the need for artificial error inflation. In fact for DA, filtering based on correction threshold would be needless as cases with low accuracy shall inevitably get down-weighted due to high uncertainty.

@@

In this study, a methodology based on quantile regression neural network (QRNN) is used for identifying and correcting the cloud contamination in operational microwave humidity channels. QRNN is a neural network which trains on all-sky brightness temperatures from channels containing orthogonal information to humidity channels, to estimate the noise free clear-sky (NFCS) brightness temperatures. The output is the posterior distribution of predictions over different quantiles. QRNN is a channel specific approach, or in other words, QRNN is trained separately for each channel, and the cloud correction for each is independent of other channels.

@@
